### 2023-07-05 paper/localpsf.tex
editing paper. about to convert\bm to \mathbf

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.34]{impulse_batch1.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.34]{impulse_batch2.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.34]{impulse_batch3.png}
	\end{subfigure}
	\caption{Three batches, $\eta_b$, of normalized impulse responses, $\impulseresponse_{x}$, that arise from applying $\Aop$ to weighted sums of scattered point sources (see Section \ref{sec:get_impulse_response}). Here $\Aop$ is the ice sheet inverse problem data misfit Gauss-Newton Hessian described in Section~\ref{sec:numerical_results}. 
	Black stars are point source locations. Shading shows the magnitude of the normalized impulse responses (darker means larger function values). Dashed gray ellipses are estimated impulse response support ellipsoids based on the moment method in Section \ref{sec:intromoments}. The large circle is $\partial \Omega$. 
	}
	\label{fig:batches_intro}
\end{figure}

@@

The proposed operator approximation method may be categorized as a PSF method that is loosely based on ``product convolution'' (PC) approximations, which are approximations of an operator by weighted sums of convolution operators with spatially varying weights. PC and PSF methods have a long history dating back several decades. We note the following papers (among many others) in which the convolution kernels are constructed from sampling impulse responses of the operator to scattered point sources:~\cite{Adorf94,AlgerEtAl19,BigotEscandeWeiss19,EscandeWeiss15,EscandeWeiss22,EscandeWeiss12,FishEtAl96,NagyOleary98,ZhuLiFomelEtAl16}. For background on PC and PSF methods, we recommend the following papers: \cite{DenisEtAl15,EscandeWeiss17,GentileCourbinMeylan13}. While PC approximations are based on an assumption of local translation invariance, the method we propose is based on a more general assumption we call ``local mean displacement invariance'' (Section~\ref{sec:local_mean_displacement_invariance} and Figure~\ref{fig:mean_displacement_invariance}). This improves the interpolation of the impulse responses. Furthermore, in our previous work~\cite{AlgerEtAl19}, we chose point sources optimally via a sequential adaptive procedure. However, in that work each point source required a separate operator application, making the previous method expensive when a large number of impulse responses is desired. In this paper, we use a new moment method (Section \ref{sec:intromoments}) which permits computation of many impulse responses (e.g., 50) per operator application. Finally, the method we propose never evaluates computed impulse responses outside of their domain of definition. This eliminates boundary artifacts which plague conventional PC methods.

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.75]{mean_displacement_invariance.pdf}
	\end{center}
	\caption{This figure illustrates impulse responses, $\impulseresponse_x$ and $\impulseresponse_{x'}$, that arise from applying $\Aop$ to point sources at two points, $x$ and $x'$. The operator $\Aop$ is locally mean displacement invariant (Section \ref{sec:local_mean_displacement_invariance}) if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(z)$ denotes the mean (center of mass) of $\phi_z$.}
	\label{fig:mean_displacement_invariance}
\end{figure}

@@

We use the proposed method to approximate the Newton (or Gauss-Newton) Hessians
in inverse problems governed by PDEs which model steady state ice
sheet flow~\cite{PetraEtAl12} (Section~\ref{sec:ice}) and transport
and flow of a contaminant~\cite{PetraStadler11}
(Section~\ref{sec:adv}). 
To reconstruct the unknown parameter fields, denoted $q$, the inverse problems are formulated as nonlinear least squares
optimization problems, whose objective functions consist of a data misfit term
(between the observations and model output) and a bi-Laplacian
regularization term following~\cite{VillaPetraGhattas21}. The
regularization is centered at a constant function $q_0(x)$. To
mitigate boundary effects
%For stokes q_0(x) = 10.5$%
we use a constant coefficient Robin boundary condition as
in~\cite{Roininen14}. The parameters for the bi-Laplacian operator are chosen
so that the Green's function of the Hessian of the regularization has
a characteristic length of $0.25$ of the domain radius. For the
specific setup, we refer the reader to~\cite[Section
  2.2]{VillaPetraGhattas21}. In all numerical results we choose the
regularization parameter (which controls the overall strength of the regularization) using the Morozov discrepancy
principle~\cite{Vogel02}.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations, $\mathbf{A}_H$, of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit). The 
approximations are indicated by ``PSF ($\nbatch$)'', where $\nbatch$ is the number of impulse response batches used to build the approximation. The Hessian of the regularization term is a combination of
stiffness and mass matrices, which are sparse. Therefore, we form
H-matrix representations of these matrices and combine them into a H-matrix approximation of the regularization term in the Hessian, $\mathbf{R}_H$, using
standard sparse H-matrix techniques. Then, H-matrix approximations of the Gauss-Newton Hessian, 
$$\bm{H}_\text{gn}\approx \widetilde{\bm{H}}:=\mathbf{A}_H + \mathbf{R}_H,$$
are formed by adding $\mathbf{A}_H$ to $\mathbf{R}_H$ using fast H-matrix arithmetic. We modify $\widetilde{\bm{H}}$ to be (approximately) symmetric positive semi-definite via the procedure described in Section~\ref{sec:flip_eigs}. We factor $\bm{\preconditioner}$
using fast H-matrix methods, then use the factorization as a
preconditioner. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel. The numerical results show that~$\widetilde{\bm{H}}$ is a good preconditioner for both~$\bm{H}_\text{gn}$ and~$\bm{H}$.

@@

\begin{figure}
	\begin{center} 
    \begin{subfigure}{0.99\textwidth}
    \begin{center} 
	\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
	\end{center}
	\caption{Ice sheet model geometry}
	\label{fig:ice_mountain_mesh}
    \end{subfigure} \\
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{mtrue_withColorBar.png}
		\caption{$\basalfriction_\text{true}$.}
		\label{fig:true_beta}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{trueVelocity2_glyphs.png}
		\caption{$\velocity_\text{true}$.}
		\label{fig:stokes_velocity}
	\end{subfigure}
    \end{center}
	\caption{(Ice sheet) 
		(\ref{fig:ice_mountain_mesh}) Bird's eye view of the ice sheet discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10^4$ meters, the maximum height of the mountain is $2.1 \times 10^3$ meters, and the average thickness of the ice sheet is 250 meters.		
		(\ref{fig:true_beta}) True parameter, $\basalfriction_\text{true}$. (\ref{fig:stokes_velocity}) True velocity, $\velocity_\text{true}$. Arrows indicate the direction of $\velocity_\text{true}$ and color indicates the magnitude of $\velocity_\text{true}$.}
	\label{fig:true_beta_u}
\end{figure}

@@

\begin{table}
	{\small
		\begin{center}
			\begingroup
			\setlength{\tabcolsep}{3pt}
			\renewcommand{\arraystretch}{1.1}
			\begin{tabular}{c| c c c | c c c | c c c}
				&  \multicolumn{3}{|c|}{PSF (5)} & \multicolumn{3}{|c|}{REG} & \multicolumn{3}{|c}{NONE} \\
				\hline
				Iter & 
				\#CG & \#Stokes & $\|\bm{g}\|$ & 
				\#CG & \#Stokes & $\|\bm{g}\|$ & 
				\#CG & \#Stokes & $\|\bm{g}\|$ \\
				0 &
				1 & 4 & 1.9e+7 &
				3 & 8 & 1.9e+7 &
				1 & 4 & 1.9e+7 \\
				1 &
				2 & 6  & 6.1e+6 &
				8 & 18 & 8.4e+6 &
				2 & 6  & 6.1e+6 \\
				2 &
				4 & 10 & 2.6e+6 &
				16 & 34 & 4.1e+6 &
				4 & 10 & 2.6e+6 \\
				3 &
				2 & 6+22 & 6.9e+5 &
				34 & 70 & 1.8e+6 &
				14 & 30 & 6.9e+5 \\
				4 &
				3 & 8 & 4.4e+4 &
				52 & 106 & 5.6e+5 &
				29 & 60 & 1.3e+5 \\
				5 &
				5 & 12 & 2.2e+3 &
				79 & 160 & 9.4e+4 &
				38 & 78 & 1.0e+4 \\
				6 &
				0 & 2 & 1.1e+1 &
				102 & 206 & 6.5e+3 &
				58 & 118 & 1.8e+2 \\
				7 &
				--- & --- & --- &
				151 & 304 & 1.2e+2 &
				0 & 2 & 5.5e-1 \\
				8 & 
				--- & --- & --- &
				0 & 2 & 2.9e-1 &
				--- & --- & --- \\
				\hline
				Total & 
				17 & 70 & --- &
				445 & 908 & --- &
				146 & 308 & --- \\
			\end{tabular}
			\endgroup
		\end{center}
	}
	\caption{(Ice sheet) Convergence history for solving the Stokes inverse problem using inexact Newton PCG. The Newton iteration is terminated when $||\mathbf{g}|| < 10^{-6} ||\mathbf{g}_0||$, where $\mathbf{g}$ is the gradient at the current iteration, and $\mathbf{g}_0$ is the gradient at the initial guess. Preconditioners shown are the proposed PSF method with five batches (PSF (5)) constructed at the third iteration, regularization preconditioning (REG), and no preconditioning (NONE). Columns titled \#CG show the number of PCG iterations used to solve the Newton system for $\bm{\widehat{\basalfriction}}$. Columns titled $\|\bm{g}\|$ show the $l^2$ norm of the gradient at $\bm{\basalfriction}$. Columns titled \#Stokes show the total number of Stokes PDE solves performed in each Newton iteration. This consists of Stokes solves performed during the line search to arrive at the current iterate from the previous iterate, plus one adjoint Stokes solve to compute the gradient at the current iterate, plus two Stokes solves per PCG iteration when solving the Newton system to compute the next search direction. Under PSF (5) and in row Iter 3, we write $6+22$ to indicate that $6$ Stokes solves were used during the standard course of the iteration, and $22$ Stokes solves were used to build the PSF (5) preconditioner.}
	\label{tab:newton_convergence_table}
\end{table}

>>>>

\begin{figure}
	\begin{subfigure}{0.49\textwidth}
		\centering
		\includegraphics[scale=0.4]{impulse_batch1.png}
		\caption{One impulse response batch}
		\label{fig:batches_intro}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\centering
		\includegraphics[scale=0.8]{mean_displacement_invariance.pdf}
		\caption{Local mean displacement invariance}
		\label{fig:mean_displacement_invariance}
	\end{subfigure}
%	\begin{subfigure}{0.32\textwidth}
%		\centering
%		\includegraphics[scale=0.34]{impulse_batch3.png}
%	\end{subfigure}
	\caption{(\ref{fig:batches_intro}) One batch, $\eta_b$, of normalized impulse responses, $\impulseresponse_{x}$, that arise from applying $\Aop$ to a weighted sum of scattered point sources (see Section \ref{sec:get_impulse_response}). Here $\Aop$ is the ice sheet inverse problem data misfit Gauss-Newton Hessian described in Section~\ref{sec:numerical_results}. 
	Black stars are point source locations. Shading shows the magnitude of the normalized impulse responses (darker means larger function values). Dashed gray ellipses are estimated impulse response support ellipsoids based on the moment method in Section \ref{sec:intromoments}. The large circle is $\partial \Omega$. (\ref{fig:mean_displacement_invariance}) Illustration of impulse responses, $\impulseresponse_x$ and $\impulseresponse_{x'}$, that arise from applying $\Aop$ to point sources at two points, $x$ and $x'$. The operator $\Aop$ is locally mean displacement invariant (Section \ref{sec:local_mean_displacement_invariance}) if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(z)$ denotes the mean (center of mass) of $\phi_z$.
	}
	\label{fig:batches_and_mdi}
\end{figure}

@@

The proposed operator approximation method may be categorized as a PSF method that is loosely based on ``product convolution'' (PC) approximations, which are approximations of an operator by weighted sums of convolution operators with spatially varying weights. PC and PSF methods have a long history dating back several decades. We note the following papers (among many others) in which the convolution kernels are constructed from sampling impulse responses of the operator to scattered point sources:~\cite{Adorf94,AlgerEtAl19,BigotEscandeWeiss19,EscandeWeiss15,EscandeWeiss22,EscandeWeiss12,FishEtAl96,NagyOleary98,ZhuLiFomelEtAl16}. For background on PC and PSF methods, we recommend the following papers: \cite{DenisEtAl15,EscandeWeiss17,GentileCourbinMeylan13}. The proposed method improves upon existing PC methods in the following ways: (1) While PC approximations are based on an assumption of local translation invariance, the method we propose is based on a more general assumption we call ``local mean displacement invariance'' (Section~\ref{sec:local_mean_displacement_invariance} and Figure~\ref{fig:mean_displacement_invariance}), which improves the interpolation of the impulse responses. (2) In our previous work~\cite{AlgerEtAl19}, we chose point sources optimally via a sequential adaptive procedure. However, in that work each point source required a separate operator application, making the previous method expensive when a large number of impulse responses is desired. In this paper, we use a new moment method (Section \ref{sec:intromoments}) which permits computation of many impulse responses (e.g., 50) per operator application. (3) The method we propose never evaluates computed impulse responses outside of their domain of definition. This eliminates boundary artifacts which plague conventional PC methods.

%\begin{figure}
%	\begin{center}
%		\includegraphics[scale=0.75]{mean_displacement_invariance.pdf}
%	\end{center}
%	\caption{This figure illustrates impulse responses, $\impulseresponse_x$ and $\impulseresponse_{x'}$, that arise from applying $\Aop$ to point sources at two points, $x$ and $x'$. The operator $\Aop$ is locally mean displacement invariant (Section \ref{sec:local_mean_displacement_invariance}) if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(z)$ denotes the mean (center of mass) of $\phi_z$.}
%	\label{fig:mean_displacement_invariance}
%\end{figure}

@@

We use the proposed method to approximate the Newton (or Gauss-Newton) Hessians
in inverse problems governed by PDEs which model steady state ice
sheet flow~\cite{PetraEtAl12} (Section~\ref{sec:ice}) and transport
and flow of a contaminant~\cite{PetraStadler11}
(Section~\ref{sec:adv}). These inverse problems are described in Sections~\ref{sec:ice} and~\ref{sec:adv}, respectively.
To reconstruct the unknown parameter fields, denoted $q$, the inverse problems are formulated as nonlinear least squares
optimization problems, whose objective functions consist of a data misfit term
(between the observations and model output) and a bi-Laplacian
regularization term following~\cite{VillaPetraGhattas21}. The
regularization is centered at a constant function $q_0(x)$. To
mitigate boundary effects
%For stokes q_0(x) = 10.5$%
we use a constant coefficient Robin boundary condition as
in~\cite{Roininen14}. The parameters for the bi-Laplacian operator are chosen
so that the Green's function of the Hessian of the regularization has
a characteristic length of $0.25$ of the domain radius. For the
specific setup, we refer the reader to~\cite[Section
  2.2]{VillaPetraGhattas21}. In all numerical results we choose the
regularization parameter (which controls the overall strength of the regularization) using the Morozov discrepancy
principle~\cite{Vogel02}.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations, $\mathbf{A}_H$, of the data misfit Gauss-Newton Hessian (the term in $\mathbf{H}_\text{gn}$ that arises from the data misfit). The 
approximations are indicated by ``PSF ($\nbatch$)'', where $\nbatch$ is the number of impulse response batches used to build the approximation. The Hessian of the regularization term is a combination of
stiffness and mass matrices, which are sparse. Therefore, we form
H-matrix representations of these matrices and combine them into a H-matrix approximation of the regularization term in the Hessian, $\mathbf{R}_H$, using
standard sparse H-matrix techniques. Then, H-matrix approximations of the Gauss-Newton Hessian, 
$$\mathbf{H}_\text{gn}\approx \widetilde{\mathbf{H}}:=\mathbf{A}_H + \mathbf{R}_H,$$
are formed by adding $\mathbf{A}_H$ to $\mathbf{R}_H$ using fast H-matrix arithmetic. We modify $\widetilde{\mathbf{H}}$ to be (approximately) symmetric positive semi-definite via the procedure described in Section~\ref{sec:flip_eigs}. We factor $\mathbf{\preconditioner}$
using fast H-matrix methods, then use the factorization as a
preconditioner. We approximate $\mathbf{H}_\text{gn}$ rather than $\mathbf{H}$ because $\mathbf{H}$ more often has negative values in its integral kernel. The numerical results show that~$\widetilde{\mathbf{H}}$ is a good preconditioner for both~$\mathbf{H}_\text{gn}$ and~$\mathbf{H}$.

@@

\begin{figure}
	\begin{center} 
    \begin{subfigure}{0.99\textwidth}
    \begin{center} 
	\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
	\end{center}
	\caption{Ice sheet model geometry}
	\label{fig:ice_mountain_mesh}
    \end{subfigure} \\
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{mtrue_withColorBar.png}
		\caption{$\basalfriction_\text{true}$}
		\label{fig:true_beta}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{trueVelocity2_glyphs.png}
		\caption{$\velocity_\text{true}$}
		\label{fig:stokes_velocity}
	\end{subfigure}
    \end{center}
	\caption{(Ice sheet) 
		(\ref{fig:ice_mountain_mesh}) Bird's eye view of the ice sheet discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10^4$ meters, the maximum height of the mountain is $2.1 \times 10^3$ meters, and the average thickness of the ice sheet is 250 meters.		
		(\ref{fig:true_beta}) True parameter, $\basalfriction_\text{true}$. (\ref{fig:stokes_velocity}) True velocity, $\velocity_\text{true}$. Arrows indicate the direction of $\velocity_\text{true}$ and color indicates the magnitude of $\velocity_\text{true}$.}
	\label{fig:true_beta_u}
\end{figure}

@@

\begin{table}
	{\small
		\begin{center}
			\begingroup
			\setlength{\tabcolsep}{3pt}
			\renewcommand{\arraystretch}{1.1}
			\begin{tabular}{c| c c c | c c c | c c c}
				&  \multicolumn{3}{|c|}{PSF (5)} & \multicolumn{3}{|c|}{REG} & \multicolumn{3}{|c}{NONE} \\
				\hline
				Iter & 
				\#CG & \#Stokes & $\|\mathbf{g}\|$ & 
				\#CG & \#Stokes & $\|\mathbf{g}\|$ & 
				\#CG & \#Stokes & $\|\mathbf{g}\|$ \\
				0 &
				1 & 4 & 1.9e+7 &
				3 & 8 & 1.9e+7 &
				1 & 4 & 1.9e+7 \\
				1 &
				2 & 6  & 6.1e+6 &
				8 & 18 & 8.4e+6 &
				2 & 6  & 6.1e+6 \\
				2 &
				4 & 10 & 2.6e+6 &
				16 & 34 & 4.1e+6 &
				4 & 10 & 2.6e+6 \\
				3 &
				2 & 6+22 & 6.9e+5 &
				34 & 70 & 1.8e+6 &
				14 & 30 & 6.9e+5 \\
				4 &
				3 & 8 & 4.4e+4 &
				52 & 106 & 5.6e+5 &
				29 & 60 & 1.3e+5 \\
				5 &
				5 & 12 & 2.2e+3 &
				79 & 160 & 9.4e+4 &
				38 & 78 & 1.0e+4 \\
				6 &
				0 & 2 & 1.1e+1 &
				102 & 206 & 6.5e+3 &
				58 & 118 & 1.8e+2 \\
				7 &
				--- & --- & --- &
				151 & 304 & 1.2e+2 &
				0 & 2 & 5.5e-1 \\
				8 & 
				--- & --- & --- &
				0 & 2 & 2.9e-1 &
				--- & --- & --- \\
				\hline
				Total & 
				17 & 70 & --- &
				445 & 908 & --- &
				146 & 308 & --- \\
			\end{tabular}
			\endgroup
		\end{center}
	}
	\caption{(Ice sheet) Convergence history for solving the Stokes inverse problem using inexact Newton PCG. The Newton iteration is terminated when $||\mathbf{g}|| < 10^{-6} ||\mathbf{g}_0||$, where $\mathbf{g}$ is the gradient at the current iteration, and $\mathbf{g}_0$ is the gradient at the initial guess. Preconditioners shown are the proposed PSF method with five batches (PSF (5)) constructed at the third iteration, regularization preconditioning (REG), and no preconditioning (NONE). Columns titled \#CG show the number of PCG iterations used to solve the Newton system for $\bm{\widehat{\basalfriction}}$. Columns titled $\|\bm{g}\|$ show the $l^2$ norm of the gradient at $\bm{\basalfriction}$. Columns titled \#Stokes show the total number of Stokes PDE solves performed in each Newton iteration. This consists of Stokes solves performed during the line search to arrive at the current iterate from the previous iterate, plus one adjoint Stokes solve to compute the gradient at the current iterate, plus two Stokes solves per PCG iteration when solving the Newton system to compute the next search direction. Under PSF (5) and in row Iter 3, we write $6+22$ to indicate that $6$ Stokes solves were used during the standard course of the iteration, and $22$ Stokes solves were used to build the PSF (5) preconditioner.}
	\label{tab:newton_convergence_table}
\end{table}
